---
title: "Week 3 — Facilitator Notes"
subtitle: "R + Quarto Orientation"
---

## Learning Objectives

Residents should leave able to create a Quarto document, write code chunks, render to HTML, and understand the relationship between source and output.

---

## Session Format

| Segment | Duration | Activity |
|---------|----------|----------|
| Demo | 20 min | Live Quarto creation and rendering |
| Guided | 25 min | Residents follow along |
| Independent | 30 min | Complete the exercise document |
| Git Integration | 15 min | Push to GitHub |

---

## Key Teaching Points

### Quarto is Literate Programming

Code and narrative in one place. This is the standard for reproducible research.

### The YAML Header Matters

Show what happens with bad YAML (indentation errors, missing quotes). Quarto fails fast with clear errors.

### Chunk Options

| Option | Effect |
|--------|--------|
| `echo: false` | Hide code, show output |
| `eval: false` | Show code, don't run |
| `warning: false` | Suppress package load warnings |

Use `echo: false` for final reports; `echo: true` for teaching.

---

## Live Demo Script

1. **File → New → Quarto Document**
   - Title: "Demo"
   - Author: Your name
   - Format: HTML

2. **Examine default template**
   - YAML at top
   - Text section
   - Code chunk

3. **Modify YAML**
   ```yaml
   ---
   title: "Drug Calculator"
   author: "Dr. Demo"
   date: "2026-04-15"
   ---
   ```

4. **Add content**
   ```markdown
   ## Patient Data
   
   Creating sample data:
   
   ```{r}
   patients <- data.frame(
     id = 1:3,
     weight = c(70, 85, 60)
   )
   patients
   ```
   
   ## Calculation
   
   Dose = weight × 2 mg/kg:
   
   ```{r}
   patients$dose <- patients$weight * 2
   patients
   ```
   ```

5. **Render** (Ctrl+Shift+K)
   - Show HTML opens automatically
   - Point out `.qmd` vs `.html` files in folder

6. **Inline code**
   - Add: `Mean dose is `r mean(patients$dose)` mg.`
   - Re-render, show inline result

---

## Common Confusion Points

| Confusion | Clarification |
|-----------|---------------|
| "Why not just use Word?" | Reproducibility. Anyone can re-run your analysis. Version control tracks changes. |
| "What's the difference between .qmd and .html?" | `.qmd` is source (you edit). `.html` is output (you share). |
| "Do I commit the HTML?" | Generally no — it's generated. Commit `.qmd` only. |
| "My chunk won't run" | Check backticks are plain ASCII, not smart quotes. |
| YAML won't parse | Indentation is 2 spaces. No tabs. |

---

## Exercise Guidance

**The exercise asks residents to:**
1. Create a Quarto doc
2. Load `tidyverse`
3. Create patient data
4. Calculate propofol dose
5. Render and push

**Watch for:**
- `library(tidyverse)` failing → guide them to `install.packages("tidyverse")` first
- Forgetting to save before rendering
- Saving in weird locations (Desktop, Downloads)

---

## Troubleshooting

| Problem | Solution |
|---------|----------|
| "Error in library(tidyverse)" | Run `install.packages("tidyverse")` |
| Render fails, no helpful message | Check YAML indentation; check for smart quotes |
| HTML doesn't open | Check if popup blocker is blocking |
| Knit button missing | Ensure file is saved with `.qmd` extension |

---

## R Basics Sidebar

If residents struggle with R syntax, use this mini-lesson:

```r
# Assignment
x <- 5          # Arrow is preferred
x = 5           # Works but less idiomatic

# Vectors
weights <- c(70, 85, 60)

# Functions
mean(weights)
```

Don't over-teach R now — Week 4 covers programming concepts in depth.

---

## Assessment

Verify each resident can:

- [ ] Create new Quarto document
- [ ] Write text, code chunk, and inline code
- [ ] Render without errors
- [ ] Explain difference between source and output
- [ ] Push `.qmd` to GitHub

---

## Preparation Checklist

- [ ] RStudio with Quarto tested
- [ ] Sample solution document prepared
- [ ] Backup plan: RStudio Cloud accounts if local install fails
